Memory for television episodes preserves event content while introducing new across-event similarities

### 2020-02-06 paper/main.tex
finished discussion revision

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/schematic}
\caption{\small \textbf{Methods overview.} We used hand-annotated descriptions of each moment of video to fit a topic model.  Three example video frames and their associated descriptions are displayed (top two rows).  Participants later recalled the video (in the third row, we show example recalls of the same three scenes from participant 13).  We used the topic model (fit to the annotations) to estimate topic vectors for each moment of video and each sentence the participants recalled.  Example topic vectors are displayed in the bottom row (blue: video annotations; green: example participant's recalls).  Three topic dimensions are shown (the highest-weighted topics for each of the three example scenes, respectively).  We also show the ten highest-weighted words for each topic.  Figure~\topics~provides a full list of the top 10 words from each of the discovered topics.}
\label{fig:schematic}
\end{figure}
% Odd name for figure in "Results" section rather than "methods" section 
% - Paxton

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The video-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across correlation values in the yellow boxes.  Their distinctiveness score was computed as the the average (over recall events) of 1 minus the average correlation between each recall event and all other recall events that do not display a box in the same row.  \textbf{B.} The (Pearson's) correlation between precision and hand-annotated memory performance. \textbf{C.} The correlation between distinctiveness and hand-annotated memory performance. \textbf{D.} The correlation between precision and the number of events recovered by the model ($k$). \textbf{E.} The correlation between distinctiveness and the number of events recovered by the model ($k$).}
  \label{fig:precision-distinctiveness}
\end{figure}

Statistical models of memory studies often treat recall success as binary (i.e., an item either was or was not recalled), or occasionally categorical \citep[e.g., to distinguish familiarity from recollection;][]{YoneEtal02}.  Such approaches are tenable in classical list-learning or recognition memory paradigms, as the presented stimuli tend to be very simple (e.g., a sequence of individual words or items).  However, the feature-rich content of a naturalistic experiences may later be described with many, highly variable levels of success.  Our framework produces a content-based model of individual stimulus and recall events by projecting the dynamic content of the video and participants' recalls into a shared topic space.  This allows for direct, quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel, \textit{continuous} metrics for analyzing naturalistic memory:~\textit{precision} and \textit{distinctiveness}.  We define precision as the ``completeness" of recall, or how fully the presented content was recapitulated in memory.  Under our framework, we quantify this for a given recall event as the correlation between the topic proportions of the recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  A second novel metric we introduce here is \textit{distinctiveness}, which we define as the ``specificity" of recall, or how unique the description of a given section of content was, compared to descriptions for other sections of content.  We quantify this for each recall event as 1 minus the average correlation between the given recall event and all other recall events not matched to the same video event.  In addition to individual events, one may also use these metrics to describe each participant's overall performance (i.e., by averaging across a participant's event-wise precision or distinctiveness scores).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, a higher precision score is correlated to both hand-annotated memory performance (Pearson's $r(15) = 0.56, p = 0.021$) and the number of recall events estimated by our model (Pearson's $r(15) = 0.85, p < 0.001$).  We also hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  We find that this distinctiveness score is related to our model's estimated number of recalled events (Pearson's $r(15) = 0.53, p = 0.028$), and while we do not find distinctiveness to be related to hand-annotated memory performance (Pearson's $r(15) = 0.28, p = 0.275$), this is not entirely surprising given how the hand-annotated memory scores were computed (see \textit{Methods}).

@@

Because they disregard presentation order-dependent variability in the stimulus content, analyses such as those in Figure \ref{fig:list-learning} enable a more sensitive analysis of presentation order-dependent temporal dynamics in free recall. Yet by the same token, they paint a wholly incomplete picture of memory for naturalistic episodes.  In an attempt to address this shortcoming, we have developed a framework in the present study that characterizes the explicit semantic content of the stimulus and subsequent recalls.  However, sensitivity to stimulus and recall content introduces a new challenge: distinguishing between levels of recall quality for a stimulus (e.g., an event) that is considered to have been ``remembered."  When modeling memory in an experimental setting, recall quality for individual events is often cast as binary (e.g., a given list item was simply either remembered or not remembered).  Various models of memory (e.g., \citealp{Yone02}) attempt to improve upon this by including confidence ratings, rendering this binary judgement instead categorical.  To better evaluate naturalistic memory quality, we introduce a continuous metric (\textit{precision}), which reflects the level of completeness of a participant's recall for a feature-rich experience.  Additionally, recall quality for a single event is typically assessed independently from that for all other events (e.g., it is difficult to ``compare" a participant's binary recall success for list item 1 to that of list item 10).  The second novel metric we introduce (\textit{distinctiveness}) is based on analyzing of the correlational structure of an individual's full set of recall events, and reflects the specificity of their memory for a single experienced event.  We find that the successful memory performance is related to 1) the precision with which the participant recounts each event and 2) the distinctiveness of each recall event (relative to the other recalled events). The first finding suggests that the information retained for \textit{any individual event} may predict the overall amount of information retained by the participant.  The second finding suggests that the ability to distinguish between temporally or semantically similar content is also related to the quantity of content recovered. Intriguingly, prior studies show that pattern separation, or the ability to discriminate between similar experiences, is impaired in many cognitive disorders as well as natural aging \citep{StarEtal10, YassEtal11c, YassStar11b}.  Future work might explore whether and how these metrics compare between cognitively impoverished groups and healthy controls.

@@

When we searched for brain structures whose responses are consistent with the video and recall topic trajectories, we identified two diffuse networks of regions \textbf{\textit{INSERT SOME BETTER COMMENTARY HERE}} (Fig.~\ref{fig:brainz}C,D).  Decoding the significance maps (Fig.~\ref{fig:neurosynth}) revealed two interesting results.  The significant

% Here's a bunch of citations to possibly add to this last sentence, though not sure if we should because the sentence is framed as a "new idea" to come out of this study:
% Takashima et al., 2007 -- vmPFC involved in stabilizing/reorganizing retrieval networks
% Tse et al., 2007; Zeithamova and Preston, 2010; Tse et al., 2011 -- vmPFC involved in integrating featurally similar experiences
% Nieuwenhuis and Takashima, 2011 -- vmPFC activity linked to evaluating memories' relevance to current situation
% Kuhl et al., 2007; Wimber et al., 2008, 2009, 2011, 2015 -- vmPFC and vlPFC inhibit contextually irrelevant memories during retrieval
% -Paxton

@@

The video was also divided by an independent coder into 50 more broad ``scenes" ``following major shifts in the narrative (e.g., location, topic, and/or time)" \citep{ChenEtal17}.  The hand-annotated memory scores for each participant we reference in our present study were generated by considering a scene to have been recalled (in a binary fashion) ``if the participant described any part of the scene."

>>>>

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/schematic}
\caption{\small \textbf{Methods overview.} We used hand-annotated descriptions of each moment of video to fit a topic model.  Three example video frames and their associated descriptions are displayed (top two rows).  Participants later recalled the video (in the third row, we show example recalls of the same three scenes from participant 13).  We used the topic model (fit to the annotations) to estimate topic vectors for each moment of video and each sentence the participants recalled.  Example topic vectors are displayed in the bottom row (blue: video annotations; green: example participant's recalls).  Three topic dimensions are shown (the highest-weighted topics for each of the three example scenes, respectively).  We also show the 10 highest-weighted words for each topic.  Figure~\topics~provides a full list of the top 10 words from each of the discovered topics.}
\label{fig:schematic}
\end{figure}
% Odd name for figure in "Results" section rather than "methods" section 
% - Paxton

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The video-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across correlation values in the yellow boxes.  Their distinctiveness score was computed as the the average (over recall events) of 1 minus the average correlation between each recall event and all other recall events that do not display a box in the same row.  \textbf{B.} The (Pearson's) correlation between precision and hand-annotated memory performance. \textbf{C.} The correlation between distinctiveness and hand-annotated memory performance. \textbf{D.} The correlation between precision and the number of video events successfully recalled, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of video events successfully recalled, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

Statistical models of memory studies often treat recall success as binary (i.e., an item either was or was not recalled), or occasionally categorical \citep[e.g., to distinguish familiarity from recollection;][]{YoneEtal02}.  Such approaches are tenable in classical list-learning or recognition memory paradigms, as the presented stimuli tend to be very simple (e.g., a sequence of individual words or items).  However, the feature-rich content of a naturalistic experiences may later be described with many, highly variable levels of success.  Our framework produces a content-based model of individual stimulus and recall events by projecting the dynamic content of the video and participants' recalls into a shared topic space.  This allows for direct, quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel, \textit{continuous} metrics for analyzing naturalistic memory:~\textit{precision} and \textit{distinctiveness}.  We define precision as the ``completeness" of recall, or how fully the presented content was recapitulated in memory.  Under our framework, we quantify this for a given recall event as the correlation between the topic proportions of the recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  A second novel metric we introduce here is \textit{distinctiveness}, which we define as the ``specificity" of recall, or how unique the description of a given section of content was, compared to descriptions for other sections of content.  We quantify this for each recall event as 1 minus the average correlation between the given recall event and all other recall events not matched to the same video event.  In addition to individual events, one may also use these metrics to describe each participant's overall performance (i.e., by averaging across a participant's event-wise precision or distinctiveness scores).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, a higher precision score is correlated to both hand-annotated memory performance (Pearson's $r(15) = 0.56, p = 0.021$) and the number of video events successfully remembered, as determined by our model (Pearson's $r(15) = 0.88, p < 0.001$).  We also hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  We find that this distinctiveness score is related to our model's estimated number of recalled events (Pearson's $r(15) = 0.55, p = 0.021$), and while we do not find distinctiveness to be related to hand-annotated memory performance (Pearson's $r(15) = 0.28, p = 0.275$), this is not entirely surprising given how the hand-annotated memory scores were computed (see \textit{Methods}).

@@

Because they disregard presentation order-dependent variability in the stimulus content, analyses such as those in Figure \ref{fig:list-learning} enable a more sensitive analysis of presentation order-dependent temporal dynamics in free recall. Yet by the same token, they paint a wholly incomplete picture of memory for naturalistic episodes.  In an attempt to address this shortcoming, we have developed a framework in the present study that characterizes the explicit semantic content of the stimulus and subsequent recalls.  However, sensitivity to stimulus and recall content introduces a new challenge: distinguishing between levels of recall quality for a stimulus (e.g., an event) that is considered to have been ``remembered."  When modeling memory in an experimental setting, recall quality for individual events is often cast as binary (e.g., a given list item was simply either remembered or not remembered).  Various models of memory (e.g., \citealp{Yone02}) attempt to improve upon this by including confidence ratings, rendering this binary judgement instead categorical.  To better evaluate naturalistic memory quality, we introduce a continuous metric (\textit{precision}), which reflects the level of completeness of a participant's recall for a feature-rich experience.  Additionally, recall quality for a single event is typically assessed independently from that for all other events (e.g., it is difficult to ``compare" a participant's binary recall success for list item 1 to that of list item 10).  The second novel metric we introduce (\textit{distinctiveness}) is based on analyzing of the correlational structure of an individual's full set of recall events, and reflects the specificity of their memory for a single experienced event.  We find that both of these metrics relate to the overall number of video events participants successfully recalled, and that our precision metric additionally relates to \cite{ChenEtal17}'s hand-annotated memory memory scores.  Though we do not find participants' average recall distinctiveness related to the hand-annotated memory scores, this is not entirely surprising given the divergence of behavior they capture.  In hand-scoring each participant's verbal recall for each of 50 (manually-deliniated) scenes, ``[a] scene was counted as recalled if the participant described any part of the scene" \citep{ChenEtal17}.  In other words, both an extensive description of a scene's content and a brief mention of some subset of its content were (binarily) considered equally successful recalls.  By contrast, we identify the event structure in participants' recalls in an unsupervised manner, independent of the video event-timeseries, prior to mapping between video and recall content.  Our HMM-based event-segmentation produces boundaries between timepoints where the topic proportions shift in a substantial way, and because a small handful of words is unlikely to contribute significantly to the topic proportions for any sliding window, such brief scene descriptions will most often not begat a sufficiently large shift in the resulting topic proportions for the HMM to identify an event boundary.  Instead, they will be grouped with a neighboring event, consequently lowering that event's distinctiveness score and by extension, the participant's overall distinctiveness score.  This is in essence the qualitative difference between distinctive and indistinctive recall, and reflects the comparison shown in Figure \ref{fig:distinctiveness-detail}C.  Intriguingly, prior studies show that pattern separation, or the ability to cleanly discriminate between similar experiences, is impaired in many cognitive disorders as well as natural aging \citep{StarEtal10, YassEtal11c, YassStar11b}.  Future work might explore whether and how these metrics compare between cognitively impoverished groups and healthy controls.

@@

In the analyses outlined in Figure \ref{fig:brainz}, we identified two diffuse networks of brain structures whose responses were consistent with the video and recall topic trajectories, respectively.  Decoding the associated significance maps with \texttt{Neurosynth} revealed two intriguing results.  First, the top 10 terms returned for the video-driven searchlight significance map were centered around themes of language and semantic meaning (Fig.~\ref{fig:neurosynth}A).  In other words, the voxels identified as more reflective of the video's temporal structure (i.e., voxels with lower permutation correction-derived $p$-values), as defined by our model, were most likely to be reported as active in studies focusd on the the neural underpinnings of semantic processing.  This finding is interesting, as our model specifically captures the temporal structure of the video's \textit{semantic} content (e.g., as opposed to that of the visual, auditory, or affective content).  This suggests that the network of structures displayed in Figure \ref{fig:brainz}C may play a roll in processing the evolving semantic structure of ongoing experiences.

Our second searchlight analysis identified a largely separate network of regions (Fig.~\ref{fig:brainz}D) whose patterns of activity as participants viewed the video reflected the idiosyncratic structure of each individual's later recall.  Decoding the associated significance map yielded a set of terms that primarily reflected names of specific structural regions (such as ``thalamus," ``anterior insula," ``anterior cingulate" and ``inferior frontal"; Fig.~\ref{fig:neurosynth}B).  Interestingly, these regions share membership in a common, large-scale functional network (termed the ``salience network") involved in detecting and processing affective cues.  In particular, the latter three regions have been implicated in functions relevant to assigning personal meaning to an experience, including: ascribing subjective value to raw, sensory input \citep{MedfCrit10}; modulating semantic and phonological processing in response to personally salient stimuli \citep{KellEtal07b}; and directing and reallocating attention and working memory resources towards the most relevant stimuli \citep{MenoUddi10}.  This suggests that the network of structures displayed in Figure \ref{fig:brainz}D may be play a roll in transforming and restructuring ongoing experiences through the lens of one's own personal values as they are encoded in memory.

@@
